---
title: "PICTographPlus"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PICTographPlus}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include = FALSE}
library(pictographPlus)
```

# 1. Introduction

PICTographPlus is a computational tool that integrates bulk DNA and RNA sequencing data to:

1) Reconstruct Clone-Specific Transcriptomic Profiles
2) Infer Tumor Evolution
3) Identify Transcriptional Transitions Between Clones

The tool infers tumor clonal evolution from single or multi-region sequencing data by modeling the uncertainty of mutation cellular fraction (MCF) in small somatic mutations (SSMs) and copy number alterations (CNAs). Using a Bayesian hierarchical model, it assigns SSMs and CNAs to subclones, reconstructing tumor evolutionary trees that adhere to principles of lineage precedence, sum condition, and optional constraints based on sample presence. For deconvolution, PICTographPlus integrates tumor clonal tree structures with clone proportions across samples to resolve bulk gene expression data. It optimizes an objective function that minimizes discrepancies between observed and predicted sample-level gene expression while imposing a smoothness penalty, ensuring that closely related clones display greater gene expression similarity. Lastly, the tool conducts pathway enrichment analysis to identify statistically significant alterations in pathways connecting tumor clones.

------

# 2. Input data file format

## 2.1 Input data (DNA) for tumor evolution reconstruction

PICTographPlus takes input data in multiple formats for flexible user inputs:

1) Three CSV files, one each for SSMs, CNAs, and germline heterozygous SNVs (section 2.1.1).
2) Two csv files, one for SSM and one for CNA (section 2.1.2).
3) A single csv file that contains SSM and CNA information (section 2.1.3).

### 2.1.1 Three CSV files, one each for SSMs, CNAs, and germline heterozygous SNVs.

When to use: if you are using a copy number caller that only outputs log2 ratio or total copy number without BAF information. e.g. cnvkit.

The SSM file should contain at least sample, mutation, total_reads, alt_reads, chrom, start, and end columns, with the purity column being optional. 

```{r}
head(read.csv(system.file('extdata/examples/example4_snv_with_purity.csv', 
                          package = 'pictographPlus')))
```
The CNA file should contain sample, chrom, start, end, and tcn columns. The total copy number (tcn) can be inferred from many copy number callers. In case a copy number caller outputs the $log2$ ratio of a segment, the tcn can be calculate using $tcn = 2 * 2^{\log_2 R}$. 

```{r}
head(read.csv(system.file('extdata/examples/example4_cna.csv', 
                          package = 'pictographPlus')))
```
The SNV file contains the count of the heterozygous germline SNVs that has the information about the "chrom", the "position" of the germline heterozygous SNV, "ref" and "alt" allele, and the reference and altenative reads counts in germline (normal) sample as well as all other samples. Note: the sample name (sample1, sample2, ... etc.) should matched the sample name used in the SSM and CNA file. 

```{r}
head(read.csv(system.file('extdata/examples/example4_SNP.csv', 
                          package = 'pictographPlus')))
```
#### Obtaining Germline Heterozygous Positions

The heterozygous germline positions can be obtained using tools such as **GATK HaplotypeCaller**. The reference and alternative read counts for each tumor sample can be retrieved using **samtools mpileup**.

We provide a Python script, `getPileUp.py`, to help generate the desired SNV file:

```bash
python getPileUp.py -v haplotype.vcf -b sample1.bam sample2.bam ... 
                    -o outputDir -f hg38.fa [--minreads] [--vaf]
```

| Parameter   | Description                                                   | Option                 |
|-------------|---------------------------------------------------------------|------------------------|
| **-v**      | VCF output from HaplotypeCaller                               | **Required**           |
| **-b**      | Tumor BAM files for tumor samples                             | **Required**           |
| **-o**      | Output directory                                             | **Required**           |
| **-f**      | Human reference genome                                       | **Required**           |
| **--minreads** | Minimum read count for both ref and alt to keep a site    | Default: 3             |
| **--vaf**   | Minimum and maximum VAF for normal heterozygous sites         | Default: 0.1 0.9       |

This script creates a folder `outputDir/pileup`, which contains a `germline_het.txt` file and all `pileup_summary.txt` files. These files are used to extract the germline heterozygous positions and convert them to the SNV file format described above. The resulting `germline_SNV.csv` can be used as input for **PICTographPlus**.

> **Note:** Ensure that the BAM file names (e.g., `sample1.bam`) match the sample names (e.g., `sample1`) used in the SSM and CNA files. If needed, rename the columns in `germline_SNV.csv` so they match the sample names in your SSM and CNA files.

### 2.1.2 Two csv files, one for SSM and one for CNA

When to use: if you are using a copy number caller that also includes BAF information. e.g. facets.

The SSM file should contain columns "sample", "mutation", "total_reads", "alt_reads", "chrom", "start", and "end". The "purity" column with be optional. 

```{r}
head(read.csv(system.file('extdata/examples/example3_snv_with_purity.csv', 
                          package = 'pictographPlus')))
```

The CNA file should contain columns "sample", "chrom", "start", "end", "tcn" and "baf". Depending on the copy number caller used, the users may need to convert the numbers to the desired format.

For example, if using facets for copy number calls, users can get tcn by using the formula $tcn = ploidy * 2^{cnlr.median.clust}$, and baf using the formula $baf = 1 / (1 + 1 / 2 ^ {mafR.clust})$

```{r}
head(read.csv(system.file('extdata/examples/example3_cna.csv', 
                          package = 'pictographPlus')))
```

### 2.1.3 A single csv file that contains SSM and CNA information

The last option is to provide a single csv file that contains at least columns named "sample", "mutation", "total_reads", "alt_reads", "tumor_integer_copy_number", and "cncf". Set cncf to 0 if a mutation has no copy number alteration. Users can also provide an optional column "major_integer_copy_number" that provides the information of the integer copy number of the major allele. If "major_integer_copy_number" is not provided, it will be estimated using an internal function built in the package. Another optional column is "purity" column that provides the information of normal contamination of a sample. 

NOTE: using this option will generate trees with SSMs only, CNA will not be assigned to clusters but only used for VAF correction.

```{r}
head(read.csv(system.file('extdata/examples/example2_snv_with_purity.csv', 
                          package = 'pictographPlus')))
```
## 2.2 Input data for bulk RNA expression

The RNA file should be a csv file of columns Gene, followed by the tumor samples (tumor sample name should match that of the genomic input), and lastly the read counts of a matched normal sample (optional). The read counts should be raw counts.

```{r}
head(read.csv(system.file('extdata/examples/rna_example.csv', 
                          package = 'pictographPlus')))
```

-----

# 3. Run PICTographPlus

## 3.1 Running PICTographPlus in one step

PICTographPlus can be run using the function `runPICTographPlus`, which runs both tumor evolution reconstruction and clone-specific transcriptomic profile deconvolution. The required files include files for genomic data and RNA expression data.

```bash
runPICTographPlus(mutation_file, copy_number_file, SNV_file, rna_file, outputDir) # if using input format 2.1.1
runPICTographPlus(mutation_file, copy_number_file, rna_file, outputDir) # if using input format 2.1.2
runPICTographPlus(mutation_file, rna_file, outputDir) # if using input format 2.1.3
```

The default deconvolution model is `tree_delta` (λ = 0.05), recommended for most use cases when a matched normal sample is available. To use a different model:

```bash
runPICTographPlus(mutation_file, rna_file, outputDir, model = "plain", lambda = 0.10)
```

## 3.2 Running PICTographPlus in multiple steps

Alternatively, the user can run tumor evolution reconstruction and bulk RNA deconvolution in separate steps.

**Tumor evolution reconstruction** can be run using:

```bash
runPictograph(mutation_file, copy_number_file, SNV_file, outputDir) # if using input format 2.1.1
runPictograph(mutation_file, copy_number_file, outputDir) # if using input format 2.1.2
runPictograph(mutation_file, outputDir) # if using input format 2.1.3
```

**Bulk RNA expression deconvolution** can be run using:

```bash
runDeconvolution(rna_file, treeFile, proportionFile, outputDir)
```
where the treeFile, and proportionFile are outputs of `runPictograph`. Users may also choose other tumor evolution reconstruction tools. File formats can be found in section 6.

The deconvolution model and regularisation strength λ can be specified:

```bash
# Default: tree_delta, lambda = 0.05
runDeconvolution(rna_file, treeFile, proportionFile, outputDir,
                 model = "tree_delta", lambda = 0.05)

# Plain Laplacian, optimised for Pearson recovery
runDeconvolution(rna_file, treeFile, proportionFile, outputDir,
                 model = "plain", lambda = 0.10)
```

**GSEA analysis using fgsea** can be run using:

```bash
X_optimal <- read.csv(paste0(outputDir, "/clonal_expression.csv"),
                      row.names=1, check.names=FALSE)

runGSEA(X_optimal, outputDir, treeFile, GSEA_file)
```

where `treeFile` is the DNA-derived clonal tree from `runPictograph` (`tree.csv`) and
`GSEA_file` is a text file of pathways of interest from [MSigDB](https://www.gsea-msigdb.org/gsea/msigdb/index.jsp).

> **Note:** GSEA always uses the DNA-derived clonal tree (`treeFile / tree.csv`) to define edges
> for pathway enrichment, even when deconvolution was run with `use_star_tree = TRUE`.
> The star topology is used only during deconvolution to avoid over-fitting to a potentially
> mis-inferred tree; once clone expressions are estimated, the true phylogeny should guide
> the edge-wise differential expression.

---------------------------------------------------

# 4. Parameters

The parameters are for function ```bash runPICTographPlus``` and the indicated module.

| Parameter | Description | default | module |
| :------------ | :----------------------------------------- | :---------- | :-------- |
| mutation_file | a csv file for SSMs | see section 2.1 | runPictograph |
| copy_number_file | a csv file for CNAs | see section 2.1 | runPictograph |
| SNV_file | a csv file for germline heterozygous SNVs| see section 2.1 | runPictograph |
| outputDir | output directory to save all outputs | NULL | runPictograph runDeconvolution runGSEA |
| sample_presence | separate mutations based on pattern of presence in all samples | TRUE | runPictograph |
| score | scoring function to estimate number of clusters. silhoutte or BIC | silhuette | runPictograph |
| max_K | maximum number of clusters | 10 | runPictograph |
| min_mutation_per_cluster | minimum number of mutations in each cluster| 5 | runPictograph |
| min_cluster_thresh | minimum MCF for each cluster | 0.05 | runPictograph |
| cluster_diff_thresh | threshold for MCF difference to merge two clusters| 0.05 | runPictograph |
| n.iter | number of iterations by JAGS | 5000 | runPictograph | 
| n.burn | number of burns by JAGS | 1000 | runPictograph |
| thin | number of thin by JAGS | 10 | runPictograph |
| mc.cores | number of cores used for parallel computing; not applicable to Windows | 8 | runPictograph |
| inits | additional parameters for JAGS | list(".RNG.name" = "base::Wichmann-Hill",".RNG.seed" = 123) | runPictograph |
| LOH | whether to include copy-neutral LOH event in CNA | FALSE | runPictograph |
| driverFile | list of driver genes used for visualization | NULL | runPictograph |
| cnv_min_length | minimum length of CNA for it to be included in the analysis | 1000000 | runPictograph |
| tcn_normal_range | range of total copy number considered as copy-neutral | c(1.75,2.3) | runPictograph |
| filter_cnv | whether or not to filter CNAs | TRUE | runPictograph |
| smooth_cnv | whether or not to process copy number alterations across samples to unify the segment start and end postions | TRUE | runPictograph |
| autosome | to only include autosome | TRUE | runPictograph |
| rna_file | bulk RNA file in integer read counts; rows are samples and columns are genes | Required | runDeconvolution |
| treeFile | tumor clone tree file | NULL | runDeconvolution runGSEA|
| proportionFile | subclone proportion file | NULL | runDeconvolution |
| normalize | normalize the raw count using DESeq2 | TRUE | runDeconvolution |
| purityFile | tumor purity file | NULL | runDeconvolution |
| lambda | regularisation strength λ for deconvolution | 0.05 | runDeconvolution |
| model | deconvolution model (see section 3.3) | "tree_delta" | runDeconvolution |
| lambda_l2 | fixed L2 Laplacian weight for elastic_net model | 0.01 | runDeconvolution |
| n_iter | IRLS iterations for adaptive/adaptive_v2 models | 5 | runDeconvolution |
| verbose | print solver convergence progress | FALSE | runDeconvolution |
| GSEA_file | geneset file in MSigDB .gmt format | NULL | runGSEA |
| top_K | top_K significant pathways to be plotted as GSEA results | 5 | runGSEA |
| n_permutations | number of permutations in fgsea | 10000 | runGSEA |
------

## 3.3 Choosing a deconvolution model

Seven deconvolution models are available. All models minimise the reconstruction
loss ‖Y − ΠX‖² with different regularisation strategies. Benchmarking was
performed on 320 synthetic pseudo-bulk replicates (P7 tumour, 4 purity levels ×
4 sample counts × 20 seeds, max tree depth = 5, K = 8 clones).

| Model | Description | Star-best λ | Recommended when |
|:------|:------------|:-----------:|:-----------------|
| `tree_delta` | Tree-delta parameterisation: X = T·Δ, group-L2 penalty on edge changes. Does not shrink true-tree LFCs. | 0.05 | **Default.** Matched normal available; best pathway F1 (synthetic edges, with-normal mode). |
| `plain` | Closed-form Laplacian smoothing. Fastest solver. | 0.10 | Highest Pearson recovery; quick exploratory runs. |
| `adaptive` | Iteratively reweighted Laplacian (IRLS). Down-weights edges with large true differences. | 0.50 | General-purpose alternative to plain. |
| `adaptive_v2` | Two-phase IRLS with unbiased initial weights (phase 1: λ=0 solve). | 0.50 | **Tumor-only mode** (no normal reference). Best F1 and sensitivity without a normal anchor. |
| `plain_debiased` | Plain Laplacian with analytic shrinkage debiasing; corrects edge LFC attenuation. | 0.50 | Matched normal; prioritise **sensitivity** (fewest missed pathways). |
| `fused_ew` | Element-wise fused LASSO via ADMM; per-gene L1 edge penalty. Requires λ > 0. | 0.01 | **Highest precision / MCC**; fewest false-positive pathways. |
| `elastic_net` | L2 Laplacian (fixed λ_L2 = 0.01) + per-gene L1 fused LASSO via ADMM. Requires λ > 0. | 0.01 | **External (population-average) normal only**; best F1 in with-extnorm mode. |

**λ selection guidance:**
- The star-best λ values above are derived from the benchmarking study and are
  topology-agnostic (selected on a star topology to avoid over-fitting to a known tree).
- If you have a matched normal sample, use `model = "tree_delta"`, `lambda = 0.05`.
- If no normal is available, use `model = "adaptive_v2"`, `lambda = 0.50`.
- `fused_ew` and `elastic_net` require `lambda > 0`; they will error at `lambda = 0`.
- The cost of using star-best λ versus true-tree-best λ is at most ≈0.008 in
  Pearson r across all models.

# 5. Output files

| File name | Description |
| :------------- | :------------- |
| mcf.csv | estimated MCF for each cluster in each sample  |
| clusterAssign.csv | cluster assignment of each SSM/CNA to each cluster  |
| CN_results.csv | estimation of the integer and major copy number of CNAs |
| tree.csv | the tree with the highest score; all trees with tied highest score is available under all_trees directory. |
| subclone_proportion.csv | estimated proportion of each cluster in each sample |
| purity.csv | estimated purity for each sample, based on the tree structure |
| tree.png | the image of a tree with the best score  |
| upsetR.png | the mutation profiles between samples; only available if number of samples is bigger than 1.|
| mcf.png | the MCF chain trace from JAGS|
| mutationClusterAssign.csv | table of all mutations information in all samples
| clone_expression.csv | clone level gene expression for each clone
| GSEA | directory contains all files from GSEA analysis

----------------------------------------------------------------

# 6. Additional file format

## 6.1 **treeFile**
A CSV file describing the tumor evolution tree used by the functions `runDeconvolution()` and `runGSEA()`.

| edge   | parent | child |
|:-------|:-------|:------|
| root->1 | root   | 1     |
| 1->2   | 1      | 2     |

## 6.2 **proportionFile**
A CSV file of subclone proportions used by `runDeconvolution()`. Each row corresponds to a clone, and each column to a sample.

|     | sample1 | sample2 |
|:----|--------:|--------:|
| **1** |   0.12  |   0.30  |
| **2** |   0.82  |   0.70  |

## 6.3 **purityFile**
A CSV file of tumor purity used by `runDeconvolution()`.

| sample1 | sample2 |
|--------:|--------:|
|    0.70 |    0.50 |

## 6.4 **driverFile**
A CSV file that contains driver mutation information used for plotting by `runPictograph()`.  
If you use this file, ensure that the corresponding mutation file follows the format **gene_extraInformation**, where **gene** is the actual gene name. The `gene_type` column can be **oncogene** or **tumor_suppressor** (leave blank if unknown).

| gene  | chrom  |    start |     end  | gene_type        |
|:------|:-------|---------:|---------:|:-----------------|
| KRAS  | chr12  |  25205246 |  25250936 | oncogene         |
| SMAD4 | chr18  |  51028528 |  51085045 | tumor_suppressor |

## 6.5 **cytobandFile**
A TSV file of cytoband information used for plotting by `runPictograph()`.

| chr  | start | end     | band   | stain  |
|:-----|------:|--------:|:-------|:-------|
| chr1 |     0 |  2300000 | p36.33 | gneg   |
| chr1 | 2300000 |  5300000 | p36.32 | gpos25 |